---
layout: post
title: "The First Stop in Monte Carlo (Methods): Rejection Sampling"
published: false
comments: true
tags: [R, "Monte Carlo"]
---

```{r setup, echo=FALSE, messagse=FALSE, results='hide'}
library(ggplot2)
set.seed(3452)
```

# Context

Monte Carlo: site of the Formula 1 Monaco Grand Prix, home to a world-famous casino, and the name of a series of interesting algorithms.

How many times have you been driving down the road thinking, "I really wish I could sample some random variables from a non-standard distribution!" Probably never.

(Application for drawing from a non-standard distribution, discrete or continuous)

## The Basics

Let's say we're interested in sampling from some completely off the wall distribution, like say this zig-zag function:

$$
z(x) = \left\{
    \begin{align}{lr}
        x & 0 \le x < 1 \\
        x - 1 & 1 \le x < 2 \\
        0 & \text{otherwise} \\
    \end{align}
    \right.
$$

```{r zig-zag}
zig.zag <- function(x) {
    if (0 <= x && x <= 1) x
    else if (1 < x && x <= 2) x - 1
    else 0
}

x <- seq(0, 2, by = 0.01)
y <- sapply(x, zig.zag)

qplot(x, y, geom = 'line')
```

Does this look even look like a probability distribution function? No, it doesn't. It's not even a continuous function, and with rejection sampling it doesn't even have to be a valid PDF! So how do we go about drawing from it, and how does it work?

### The 10,000 feet high perspective

To start off, we draw a box that completely surrounds our zig-zag function. Then, we pick a point at random from our box. If it falls under our zig-zag function, we say we accept this point and return it in the sample. If it falls above the zig-zag function, we reject it and repeat the drawing process. Do this again and again until you get the number of samples that you want. It's a very simple algorithm.

Let's look at an example outcome of these drawings.

```{r}
# We want 5000 samples from our zig-zag distribution
N <- 5000
i <- 1
draws <- c()
ret <- c()

while (i <= N) {
    # Draw a number along the width of the box
    # (the support of zig-zag)
    T <- runif(1, min=0, max=2)
    # Draw a number along the height of the box
    U <- runif(1, min=0, max=1)

    draws <- rbind(draws, c(T, U))

    # If the point is under the zig-zag,
    # accept it
    if (U <= zig.zag(T)) {
        ret <- c(ret, T)
        i <- i + 1
    }
    # Otherwise, try again
}

summary(ret)
```

A plot of our data drawn samples and wether or not they were accepted:

```{r plot, echo=FALSE}
draws <- as.data.frame(draws)
draws[,3] <- factor(draws[,2] <= sapply(draws[,1], zig.zag),
                    levels=c(TRUE, FALSE),
                    labels=c("Accept", "Reject"))
names(draws) <- c("T", "U", "Sample")

ggplot() +
  geom_line(aes(x, y)) +
  geom_point(aes(x=T, y=U, color=Sample), data=draws)
```

A histogram of our accepted data points:

```{r, echo=FALSE}
qplot(ret, binwidth=2.0/50.0) + xlab("Value") + ylab("Count")
```

That's pretty close to our distribution. It looks like we're sampling from our zig-zag!

## Down the Rabbit Hole (Diving Into the Details)

Now, I'm going to dive down into the math behind this. I encourage you to keep reading, but if you're just looking for an intuitive feel for this, you can stop here. The above is sufficient for the general idea.

The entire idea of rejection sampling hinges on the idea that if we want to sample from $$f(t)$$ can find an easy to sample density, $$d(t)$$, and a number, $$M$$, such that $$f(t) \le M d(t)$$ for every $$t$$, then 

$$
p(X) = p(T\"|\"Accept)
$$

### Great. What's that mean, and can you prove it?



# Your Turn

Now that you can sample from whatever distribution you want. Can you think of a function you'd want to sample from? What if you wanted to sample from an observed sample distribution? (Those are called empirical density functions.) That's easy now!

As always, if you have any questions or comments, commend below or write me an [e-mail](mailto:thexbarblog@gmail.com).